\documentclass[12pt,letterpaper,twoside,openright]{report}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage[backend=biber, style=nature, sorting=none]{biblatex}
\usepackage{chngcntr}
\usepackage{float}
\usepackage[top=1in, bottom=1.5in, left=1.25in, right=1in]{geometry}
\usepackage{graphicx}
\usepackage[pdftex,pdfpagelabels,bookmarks,hyperindex,hyperfigures]{hyperref}
\usepackage{mathrsfs}
\usepackage{mathtools}
\usepackage[version=3]{mhchem}
\usepackage{nicefrac}
\usepackage{setspace}
\usepackage[textsize=footnotesize,textwidth=0.75in]{todonotes}

\bibliography{biblio}
\counterwithin{equation}{section}
\counterwithin{figure}{section}
\graphicspath{{img/}}
\renewcommand\baselinestretch{1.5}

\begin{document}
\begin{titlepage}
        \vspace*{1.5cm}
	\centering \large
	{\Large \bf Stochastic models of genetic circuits\par}
        \vspace{0.3cm}
        {by\par}
        \vspace{0.3cm}
	{\Large Luis Alberto Guti\'errez L\'opez \par}
        \vspace{2.2cm}
        {\setstretch{1.0}
	{Submitted to the Department of Physics\par}
        {in partial fulfillment of the requirements for the degree of \par}}
        \vspace{.25cm}
	{Bachelor of Science in Physics\par}
        \vspace{.25cm}
        {at the\par}
        \vspace{.25cm}
        {UNIVERSIDAD DE LOS ANDES\par}
        \vspace{2.2cm}
        {Supervisor: Juan Manuel Pedraza Leal\par}
        \vspace{2.8cm}
        {May 12, 2016}

        
        
        
        
        %Author \dotfill \par
        %\setstretch{1.0}\raggedleft Department of Physics\\May 12, 2016    
        %\vspace{1.2cm}
        %\vspace{3.0cm}

        %Certified by \dotfill \par
        %\setstretch{1.0}\raggedleft Juan Manuel Pedraza Leal \\ Associate Professor \\ Monograph Supervisor
	\vfill
\end{titlepage}

\cleardoublepage

\begin{center}
{\large \bf Stochastic models of genetic circuits} \\
%by \\
%{\large Luis Alberto Guti\'errez L\'opez}
\end{center}
%\vspace{1.0cm}
%\setstretch{1.0}{
%Submitted to the Department of Physics \\
%on May 12, 2016 in partial fulfillment of the \\
%requirements for the degree of \\
%Bachelor of Science in Physics
%}
%\end{center}
\par
\subsection*{Abstract}
\addcontentsline{toc}{chapter}{Abstract}
\vspace{-0.5cm}
\begin{singlespacing}
All living beings store their genetic information in the DNA and use similar mechanisms to read it and, according to its sequence, build their structure and develop their functions. Nevertheless, the information codified in the DNA is not the only aspect that makes an organism what it is. Genes interact between them and with environmental signals through a complex network of gene regulation that determines which genes are read at a particular moment and the intensity of their activity. This makes possible, for instance, to differentiate between our muscle cells and our neurons, cells that in spite of having the same genetic material, are very different in form and function. Since gene expression is mediated by chemical reactions, which at microscopic levels are random collisions of individual molecules of the reactants, its regulation is subjected to noise. Cells also regulate the expression of many genes according to the environment, which may change randomly. In response to these important sources of noise, living beings have developed their regulatory networks to work properly under its presence. This work explores some of the models that have been done on the last years related to stochasticity in gene expression, and the insights they have given to us into the principles of biology and the design of synthetic biological circuits.\par
\end{singlespacing}
\subsection*{Resumen}
\vspace{-0.5cm}
\begin{singlespacing}
Todos los seres vivos almacenan su informaci\'on gen\'etica en el ADN y usan mecanismos similares para leerla y, de acuerdo a su secuencia, construir su estructura f\'isica y desarrollar sus funciones. Sin embargo, la informaci\'on codificada en el ADN no es el \'unico aspecto que determina a un organismo. Los genes interact\'uan entre ellos y con se\~nales externas a trav\'es de una compleja red de regulaci\'on gen\'etica que determina cu\'ales genes son son le\'idos en cierto momento particular y la intensidad de su actividad. Esto permite, por ejemplo, diferenciar entre nuestras c\'elulas musculares y nuestras neuronas, c\'elulas que a pesar de que poseen el mismo material gen\'etico, son muy diferentes en forma y funci\'on. Debido a que la expresi\'on gen\'etica es mediada por reacciones qu\'imicas, las cuales a niveles microsc\'opicos son colisiones aleatorias entre mol\'eculas individuales de los reactantes, existe ruido en su regulaci\'on. Las c\'elulas tambi\'en regulan la expresi\'on de muchos de sus genes de acuerdo a su ambiente, el cual puede cambiar aleatoriamente. En respuesta a estas importantes fuentes de ruido, los seres vivos han desarrollado sus redes gen\'eticas para que funcionen adecuadamente bajo su presencia. Este trabajo explora algunos de los modelos que han sido realizados en los \'ultimos a\~nos relacionados con estocasticidad en la expresi\'on gen\'etica, y las conclusiones a las que nos han permitido llegar acerca de los principios de la biolog\'ia y el dise\~no de circuitos biol\'ogicos sint\'eticos.\par
\end{singlespacing}
\setstretch{1.5}
\newpage
\tableofcontents
\newpage
\listoffigures
%\newpage
%\listoftables

\raggedbottom
\include{introduction}
\include{concepts}
\include{master1}
\include{fdt}
\include{langevin1}
\include{bursting}
\include{division}
\include{final}
\printbibliography
\addcontentsline{toc}{chapter}{Bibliography}

\end{document}
